Spherical harmonic analysis of faceted spheroids identifies shaping strategies and standardisation at Qianshangying (North China)

### 2025-09-04 analysis/paper/paper.qmd
updates of texts

Among the shape descriptors of spheroids, we observed several significant correlations. Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with Spherical Harmonics Power (SHE) (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms. As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids [@tittonSubspheroidsLithicAssemblage2020]. This correlation between edge angle and sphericity may help explain why spheroids were often interpreted as exhausted cores in earlier research [@sahnouni1997]. Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult. However, the reverse is not necessarily true, as not all exhausted cores evolve into spheroids.

@@

However, no significant correlations were observed between reduction variables and sphericity or curvature. This implies that increased reduction does not necessarily produce smoother surfaces or perfect spherical surfaces. Such patterns likely reflect the technological nature of the sample: the analyzed spheroids and sub-spheroids are "faceted" products of flake removal. Flaking increases surface complexity by adding edges and angular intersections, which may elevate curvature while also expanding surface area, factors that can distort sphericity values. For instance, a smooth, naturally rounded cobble might exhibit lower curvature and a sphericity closer to 1 than a heavily flaked spheroid, even if both appear similarly rounded overall. Therefore, the negative correlation between reduction intensity and SHE, but not sphericity, suggests that shaping efforts prioritized achieving symmetrical, coherent forms rather than naturally polished, smooth surfaces.

@@

The evolutionary origins of shaping strategies are most commonly linked to the Acheulean, particularly those exemplified by bifacial shaping strategies on handaxes [@isaac1986foundation]. These strategies gradually became more refined, with mature bifacial shaping marked by the concept of bifacial symmetry, thinning and edge refinement, and the achievement of a standardized final shape [@beyene2013]. Compared to flaking technology in Oldowan, these processes demand greater cognitive planning and precise manual control [@wynn1995; @stout2015]. Although the spheroids from Qianshangying do not qualify as Large Cutting Tools (LCTs), nor does the site yield other typically Acheulean products, the faceted spheroids nonetheless exhibit a level of technological and morphological complexity comparable to that of LCTs. They also share key features, including a hierarchical and standardized reduction sequence and a high degree of symmetry and surface regularity in the final product. <!-- BM: check Shipton, see if he has anything relevant to mention here -->

@@

## Are the Qianshangying spheroids typical of East Asian assemblages?

@@

At DC and MJQ, thick, flat cobbles were selected and shaped through peripheral flaking, producing roughly approximately cubic shape [@Wei1984; @Tao1984Dingcun]. MJQ specimens retain parts of the cortex, while at DC, flakes with large exterior and small interior platform angles were removed from blanks to increase edge angles and enhance sphericity [@Tao1984Dingcun; @Wei1984]. Of the six sites exhibiting similar strategies to Qianshangying spehroids, all but MJQ are located within the Fen River and Lishui River regions. These sites were grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by trihedral picks [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe]. The trihedral picks were produced by striking the dorsal surfaces of large flakes to concentrate edges at the tip, resulting in triangular outlines and robust pointed ends, classifying them as large cutting tools (LCTs)[kuman2014large; li2018currently; yang2016late]. This is a significant observation, as it implies that spheroid production techniques similar to those at Qianshangying are more frequently associated with lithic assemblages characterized by intentional shaping strategies.

It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Acheulean technology. We infer that faceted spheroids in China appear to represent a shaping tradition that shares the same core technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], and this explains why they often occur in Acheulean techno-complex. We now find that faceted spheroids appear in small flake tool industry in northern China, whereas it was previously considered part of a simple core and flake technology, then this would suggest that small flake tool industry is more complex than previously expected.

This pattern aligns with spheroid findings in other regions of the world. Early examples of spheroids have been identified in various lithic assemblages across Africa, Europe, the Levant, and Asia, where the associated techno-complexes are generally attributed to either Oldowan, Acheulean or MP/MSA. Spheroids appear in what is considered the Late Oldowan, if a chronological boundary is set at 2.0 Ma to distinguish between Early and Late Oldowan [@gallotti2018emergence]. The growing recognition of technological diversity within the Oldowan has led to a broader consensus that the Late Oldowan includes signs of innovation, such as more complex flaking patterns and decreased morphological variability in flakes [@gallotti2018emergence; @braun2019]. Spheroids, we propose here, may represent one of the key indicators of this increasing technological complexity during the Late Oldowan [@tittonSubspheroidsLithicAssemblage2020].

In particular, faceted spheroids, which we hypothesize represent a deliberate shaping strategy, may indicate a level of technical skill approaching that of the earliest Acheulean, or perhaps even foreshadowing their emergence. After all, faceted spheroids already exhibit one of the innovations characteristic of the Acheulean—namely, the application of a specific mental template onto blanks to produce tools with broadly similar overall forms [isaac1986foundation; de2016origins].

@@

>>>>

Among the shape descriptors of spheroids, we observed several significant correlations. Mean edge angle was positively correlated with sphericity (ρ = `r cor_angle_sph`, p = $`r p_angle_sph`$) and negatively correlated with Spherical Harmonics Power (SHE) (ρ = `r cor_angle_she`, p = $`r p_angle_she`$), indicating that greater edge angles are associated with more spherical and regular forms. As noted in previous studies, mean edge angle is a key variable in capturing the morphology of spheroids [@tittonSubspheroidsLithicAssemblage2020]. This correlation between edge angle and SHE may help explain why spheroids were often interpreted as exhausted cores in earlier research [@sahnouni1997]. Indeed, spheroids must be significantly reduced, as achieving a near-perfect spherical shape typically results in increased edge angles, making further flake removal difficult. However, the reverse is not necessarily true, as not all exhausted cores evolve into spheroids.

@@

However, no significant correlations were observed between reduction variables and sphericity or curvature. This implies that increased reduction does not necessarily produce smoother surfaces or perfect spherical surfaces. Such patterns likely reflect the technological nature of the sample: the analyzed spheroids and sub-spheroids are "faceted" products of flake removal. Flaking increases surface complexity by adding edges and angular intersections, which may elevate curvature while also expanding surface area, factors that can distort sphericity values. For instance, a smooth, naturally rounded cobble might exhibit lower curvature and a sphericity closer to 1 than a heavily flaked spheroid, even if both appear similarly rounded overall. Therefore, the negative correlation between reduction intensity and SHE and mean edge angle, but not sphericity or curvanture, suggests that shaping efforts prioritized achieving symmetrical, coherent forms rather than naturally polished, smooth surfaces.

@@

The evolutionary origins of shaping strategies are most commonly linked to the Acheulean, particularly those exemplified by bifacial shaping strategies on handaxes [@isaac1986foundation]. These strategies gradually became more refined, with mature bifacial shaping marked by the concept of bifacial symmetry, thinning and edge refinement, and the achievement of a standardized final shape [@beyene2013; @shipton2019were]. Compared to flaking technology in Oldowan, these processes demand greater cognitive planning and precise manual control [@wynn1995; @stout2015]. Although the spheroids from Qianshangying do not qualify as Large Cutting Tools (LCTs), nor does the site yield other typically Acheulean products, the faceted spheroids nonetheless exhibit a level of technological and morphological complexity comparable to that of LCTs. They also share key features, including a hierarchical and standardized reduction sequence and a high degree of symmetry and surface regularity in the final product.

@@

## QSY spheroids in the context of East Asia

@@

We found six sites exhibit similar strategies to QSY. For example, at DC and MJQ, thick, flat cobbles were selected and shaped through peripheral flaking, producing roughly approximately cubic shape [@Wei1984; @Tao1984Dingcun]. MJQ specimens retain parts of the cortex, while at DC, flakes with large exterior and small interior platform angles were removed from blanks to increase edge angles and enhance sphericity [@Tao1984Dingcun; @Wei1984]. all Of the six sites exhibiting similar strategies to Qianshangying spehroids, but MJQ are located within the Fen River and Lishui River regions. These sites were grouped into the Keheng–Dingcun techno-complexes and Lishui techno-complexes, characterized by trihedral picks [@LixianMuseum1992; @Tan1999; @Fang2002; @SXA2014Dingcun; @Jia1962Kehe]. The trihedral picks were produced by striking the dorsal surfaces of large flakes to concentrate edges at the tip, resulting in triangular outlines and robust pointed ends, classifying them as large cutting tools (LCTs)[kuman2014large; li2018currently; yang2016late]. This is a significant observation, as it implies that spheroid production techniques similar to those at Qianshangying are more frequently associated with lithic assemblages characterized by intentional shaping strategies.

It should be emphasized, however, that we do not intend to imply a direct association between faceted spheroids and Acheulean technology. We infer that faceted spheroids in China appear to represent a shaping tradition that shares the same core technological competencies as LCTs [@leakey1971; @Cabanes2024; @isaac1986foundation], and this explains why they often occur in Acheulean techno-complex. We now find that faceted spheroids occur at QSY within the small flake tool industry of northern China. Since this industry was previously regarded as a simple core and flake technology, the presence of faceted spheroids suggests that it is more complex and exhibits greater variability than previously thought.

Similarly, we may need to reconsider the technological complexity implied by faceted spheroids, even when they occur in early cultural contexts traditionally regarded as simple. Faceted spheroids appear in what is considered the Late Oldowan, if a chronological boundary is set at 2.0 Ma to distinguish between Early and Late Oldowan (Gallotti et al., 2018; Leakey, 1921). The growing recognition of technological diversity within the Oldowan has led to a broader consensus that the Late Oldowan includes signs of innovation, such as more complex flaking patterns and decreased morphological variability in flakes (Braun et al., 2019; Gallotti et al., 2018). Faceted spheroids, we propose here, may represent one of the key indicators of this increasing technological complexity during the Late Oldowan (Titton et al., 2020).

In particular, if future research can demonstrate that these faceted spheroids also represent a deliberate shaping strategy, this may indicate a level of technical skill approaching that of the earliest Acheulean, or perhaps even foreshadowing its emergence. After all, faceted spheroids already embody one of the key innovation characteristics of the Acheulean, that is, the imposition of a specific mental template onto blanks in order to produce tools with broadly similar overall forms [isaac1986foundation; de2016origins].

@@

In principle, any closed three-dimensional object can be analyzed using spherical harmonics. In archaeology, materials such as spheroids or cores, ceramic vessels, plant seeds, and animal crania are particularly well suited to this approach, as their overall forms approximate spheres or ellipsoids, making them readily mappable onto the unit sphere for spherical harmonic expansion. Low-order expansions efficiently capture their primary morphological features, while the combination of low-order and high-order coefficients allows for a distinction between overall shape (e.g., symmetry, inflation, elongation) and local details (e.g., decorations, ridges, protrusions). Moreover, these objects generally do not emphasize orientation, enabling the use of rotationally invariant spherical harmonic power spectra to quantify shape variation. For objects with weaker symmetry or greater surface complexity, however, higher-order expansions are required, increasing computational cost and complicating the interpretation of coefficients.
